Towards a standardized evaluation of imputation methodology

### 2022-03-29 Formatted/Formatted.tex
Write 'obtain truth' subsection

After deciding upon a data generating mechanism, another influential--but often overlooked--decision awaits. Simulation studies on missing data methodolology have the unique option to exclude sampling variance from their evaluations, and only use the missing data generation procedure as source of Monte Carlo variation. After all, we are interested in the missingness and are not considering the noise induced by the sampling mechanism for evaluation in such studies. Therefore, it is not necessary to take sampling variation into account in order to obtain information about a method's ability to handle the missing data problem, or to objectively compare methods on their ability to correct for missingness \citep[see for a detailed discussion][]{vink14}. A lot of published work does not consider this option, while it could have sharpened inconclusive or obfuscated results. The only required change to simulation set-ups is to remove sampling variance from performance evaluation. We highly value the additional flexibility that such a simulation set-up offers. Especially in the case of data transformations it can be challenging to derive the true parametric references to evaluate against. Simply using a single generated complete set in which simulated missingness is induced and evaluating against that generated set awards a plethora of procedural problems and computational challenges.

@@

Evaluating imputation methodology requires a missing data problem to be solved. After establishing a comparative truth from a data generating mechanism, some form of missingness therefore has to be induced. Often, however, reports of simulation studies remain vague about the actual missingness conditions under investigation and, even worse, some authors only report something like
\begin{quote}
\textit{We generated missing data following a missing at random missingness mechanism.}
\end{quote}
This should be considered unacceptable as claims about the validity of the imputation inference heavily depend on the simulated missingness conditions, such as missingness patterns and missingness mechanisms. Missingness patterns concern the location of missing entries across incomplete data, whereas missingness mechanisms describe the relationship between missing entries and observed data values \citep[][p. 8]{litt20}. Under this definition, there is a row-wise element to the missingness pattern describing which variables are jointly observed, and a column-wise element encompassing the proportion of missingness in the data.

@@

\begin{table}[htb]
\begin{center}
\caption{Missingness mechanisms. [Add PMAR]}
\begin{tabular}{lll}
\hline
Mech. & Interpretation and consequences of the mechanism \\
\hline  
MCAR  & The probability to be missing is the same for all cases. In other words, the missing \\
      & values are missing at random and the observed values are observed at random. \\
MAR   & The probability to be missing is the same within groups of cases defined by the \\
      & observed data only. The essence is that observed relations in the data inform the \\
      & missingness, such that MCAR may be assumed within the observed groups. \\
MNAR  & The probability to be missing depends on unobserved aspects of the data. The \\
      & cause of the missingness is unknown and cannot be inferred from the observed data. \\
      & This is considered non-ignorable missingness \citep[see e.g.][]{rubi76}. \\
\hline
\end{tabular}
\end{center}
\end{table}

@@

[Missingness type is often ignored (text duplicated from solutions section):] Given the simulated data distributions, one random missingness model may be far more disastrous to the observed information than another model. This may influence the performance of some (but not necessarily all) imputation routines. For example, inference from hot-deck techniques such as predictive mean matching \citep{little1988missing, rubin1986statistical} may be more severely impacted by large amounts of one-tailed missingness than inference from parametric techniques. It would be a shame to overlook such results due to the focus on a single MAR mechanism.

@@

Standardizing the evaluation of imputation methodology requires simulators--including ourselves--to consider a couple of key aspects in their simulation workflows. Table \ref{table:steps} outlines some suggested steps to adopt. Please note that there is--of course--no 'silver bullet' and that we do not claim to present a universally applicable approach. Our suggestions are the accumulated result of scientific literature, research experience, and discussion. For an excellent overview of general best practices for method evaluation by means of simulation, see \citet{morr18}. We recommend adhering to their proposed ADEMP structure (aims, data-generating mechanisms, estimands, methods, performance measures) for planning and reporting simulation studies, but highlight and add some specific recommendations for the evaluation of imputation methodology.

@@

Before setting up their simulation study, the simulator should clearly define the scope of their evaluations. Evaluating imputation methodology may entail e.g. comparisons between several imputation methods, investigations of the validity of a single (novel) method, or an application to varying characteristics of a specific use case. The choice of imputation method(s) under evaluation and the simulation study's aims are highly intertwined. 
[Add that some simulation studies do not use a `ground truth' at all, they just look for the method with the best predictive performance for a certain completely observed target variable. These studies apply different combinations of imputation and estimation methods on one or more benchmark data sets, and assess the RMSE of the predicted values after imputation and estimation to say something about the comparative performance of the methods. Refer to missing data chapter \citep{liu21}. In this context, there is no talk of retrieving the `true' but unobserved values at all. The DGM is by definition design-based and there is no sampling variance. The missingness generation step is skipped and the only performance measure that can be used between methods is RMSE and maybe comparison with repect to CCA in terms of bias.]

@@

[Decide which data generation mechanism to use and obtain true value of the estimand(s).]

[Choose sampling variance:] The simulator must decide whether there is a need for sampling variance in the simulations scheme. If sampling variance is not of interest, the procedure outlined in section XYZ can be used to simulate missingness in a single complete data set. 
% For situations where sampling variance is not of interest, the sampling process can be omitted and a single complete dataset can simply be obtained from the data generating mechanism. 
The parameters of the complete sample will serve as comparative truth during the simulation. This process is computationally convenient because only a single complete data set has to be considered during all of the simulations. [insert part about transformed data here?]. When deviating from the standard simulation workflow, conventional pooling rules for multiple imputation \citep[cf.][p. 76-77]{rubi87} do not apply. Instead, alternative pooling rules need to be used \citep{raghunathan2003multiple,vink14}. Table \ref{table:dgm_truth} displays the source of the comparative truth in simulation studies with and without sampling variance. If sampling variance cannot be omitted from the simulation scheme, draw data repeatedly and apply conventional pooling rules \citep[cf.][p. 76-77]{rubi87}.

[Choose DGM:] The design-based approach is often used in situations where a probability distribution is not available, or where real-life data structures are of interest. Applications of design-based simulation are often found in official statistics. A benefit of design-based simulation is the ability to use real-life observed data structures. For evaluations of model-based simulations, it could be convincing to demonstrate a method's applicability to real-world missing data problems. This can, for example, be achieved by obtaining and imputing a fully observed set of variables, wherein the missingness is mimicked from a similar, incomplete set. Alternatively, an incomplete set could be obtained and truth could be established by removing the incomplete cases from the data. However, the real-world missingness would then be omitted.

[Obtain truth: In model-based simulation data are drawn from a known probability distribution, so the theoretical parameters under which the samples are obtained serve as the comparative truth in the simulations. In design-based simulation, data are sampled from a sufficiently large set, so the parameters of this specific set serve as the comparative truth in the simulations. But this is not a complete picture, see Table \ref{dgm_truth}.]

\begin{table}[htb]
\begin{center}
\caption{Comparative truth in model-based and design-based simulation.}
\label{table:dgm_truth}
\begin{tabular}{lll}
\hline
               & With sampling variance      & Without sampling variance \\
               & (multiple samples drawn)    & (one sample drawn) \\
\hline  
Model-based    & Truth from model            & Truth from sample \\
Design-based   & Truth from large data set   & Truth from sample \\
\hline
\end{tabular}
\end{center}
\end{table}

@@

[Missingness mechanisms:] 
First, one should always consider MCAR missingness, i.e. the scenario where the missing values are missing at random and the observed values are observed at random. % Under MCAR, the statistical properties of the observed data given the missing data are known and any imputation routine that cannot at least mimic the performance of the observed data inference, should be deemed inefficient in the scope of the simulation. 
Next, missing data should be induced conform a model that is dependent on the observed data. A straightforward technique for inducing different forms of univariate MAR missingness is described in \citet[][p. 63]{fimd} and a generalization to multivariate MAR missingness can be found in \citet[][Appendix B]{buur06} and \citet[][\S 5.2.3]{brand1999development}. If the missingness is to be induced in longitudinal data, autoregressive MAR models \citep[e.g. cf.][model 2 and model 3]{shara2015randomly} can be useful.
[Although often ignored, it can be argued that MNAR is the more likely mechanism for real-life missingness scenarios. It would be wise to include the mechanism in your simulation study, or perform sensitivity analyses.] An indication of the validity of the obtained inference, given that the assumed missingness mechanism is suspected to be invalid, may be obtained by performing sensitivity analysis \citep[see e.g.][part 5]{molenberghs2014handbook}.

[Missingness type:] 
Third, it is advisable to investigate varying shapes of MAR missingness to achieve a more realistic indication of the robustness of the imputation performance across the range of random missingness. % Given the simulated data distributions, one random missingness model may be far more disastrous to the observed information than another model. This may influence the performance of some (but not necessarily all) imputation routines. For example, inference from hot-deck techniques such as predictive mean matching \citep{little1988missing, rubin1986statistical} may be more severely impacted by large amounts of one-tailed missingness than inference from parametric techniques.
It would be a shame to overlook results due to the focus on a single MAR mechanism.

@@

>>>>

After deciding upon a data generating mechanism, another influential--but often overlooked--decision awaits. Simulation studies on missing data methodolology have the unique option to exclude sampling variance from their evaluations, and only use the missing data generation procedure as source of Monte Carlo variation. After all, we are interested in the missingness and are not considering the noise induced by the sampling mechanism for evaluation in such studies. Therefore, it is not necessary to take sampling variation into account in order to obtain information about a method's ability to handle the missing data problem, or to objectively compare methods on their ability to correct for missingness \citep[see for a detailed discussion][]{vink14}. A lot of published work does not consider this option, while it could have sharpened inconclusive or obfuscated results. The only required change to simulation set-ups is to remove sampling variance from performance evaluation. We highly value the additional flexibility that such a simulation set-up offers. Especially in the case of data transformations it can be challenging to derive the true parametric references to evaluate against. Simply using a single generated complete set in which simulated missingness is induced and evaluating against that generated data set avoids a plethora of procedural problems and computational challenges.

@@

Evaluating imputation methodology requires a missing data problem to be solved. After establishing a comparative truth from a data generating mechanism, some form of missingness therefore has to be induced. Often, however, reports of simulation studies remain vague about the actual missingness conditions under investigation and, even worse, some authors only report something like
\begin{quote}
\textit{We generated missing data following a missing at random missingness mechanism.}
\end{quote}
This should be considered unacceptable as claims about the validity of the imputation inference heavily depend on the simulated missingness conditions, such as missingness mechanisms and missingness patterns. Missingness mechanisms describe the relationship between missing entries and observed data values \citep[generally categorized into MCAR, MAR and MNAR mechanisms cf][see Table \ref{table:mech}]{rubi76}.  Missingness patterns concern the location of missing entries across incomplete data \citep[][p. 8]{litt20}. Under this definition, there is a row-wise element to the missingness pattern describing which variables are jointly observed, and a column-wise element encompassing the proportion of missingness in the data.

@@

\begin{table}[htb]
\begin{center}
\caption{Missingness mechanisms. [Add PMAR]}
\label{table:mech}
\begin{tabular}{lll}
\hline
Mech. & Interpretation and consequences of the mechanism \\
\hline  
MCAR  & The probability to be missing is the same for all cases. In other words, the missing \\
      & values are missing at random and the observed values are observed at random. \\
MAR   & The probability to be missing is the same within groups of cases defined by the \\
      & observed data only. The essence is that observed relations in the data inform the \\
      & missingness, such that MCAR may be assumed within the observed groups. \\
MNAR  & The probability to be missing depends on unobserved aspects of the data. The \\
      & cause of the missingness is unknown and cannot be inferred from the observed data. \\
      & This is considered non-ignorable missingness \citep[see e.g.][]{rubi76}. \\
\hline
\end{tabular}
\end{center}
\end{table}

@@

[Missingness type is often ignored:] Given the simulated data distributions, one random missingness model may be far more disastrous to the observed information than another model. This may influence the performance of some (but not necessarily all) imputation routines. For example, inference from hot-deck techniques such as predictive mean matching \citep{little1988missing, rubin1986statistical} may be more severely impacted by large amounts of one-tailed missingness than inference from parametric techniques. It would be a shame to overlook such results due to the focus on a single MAR mechanism.

@@

Standardizing the evaluation of imputation methodology requires simulators--including ourselves--to consider a couple of key aspects in their simulation workflows. Table \ref{table:steps} outlines some suggested steps to adopt. Please note that there is--of course--no `silver bullet' and that we do not claim to present a universally applicable approach. Our suggestions are the accumulated result of scientific literature, research experience, and discussion. For an excellent overview of general best practices for method evaluation by means of simulation, see \citet{morr18}. We recommend adhering to their proposed ADEMP structure (aims, data-generating mechanisms, estimands, methods, performance measures) for planning and reporting simulation studies, but highlight and add some specific recommendations for the evaluation of imputation methodology.

@@

Before setting up their simulation study, the simulator should clearly define the scope of their evaluations. Evaluating imputation methodology may entail e.g. comparisons between several imputation methods, investigations of the validity of a single (novel) method, or an application to varying characteristics of a specific use case. The choice of imputation method(s) under evaluation and the simulation study's aims are highly intertwined.

[Add that some simulation studies do not use a `ground truth' at all, they just look for the method with the best predictive performance for a certain completely observed target variable. These studies apply different combinations of imputation and estimation methods on one or more benchmark data sets, and assess the RMSE of the predicted values after imputation and estimation to say something about the comparative performance of the methods. Refer to missing data chapter \citep{liu21}. In this context, there is no talk of retrieving the `true' but unobserved values at all. The DGM is by definition design-based and there is no sampling variance. The missingness generation step is skipped and the only performance measure that can be used between methods is RMSE and maybe comparison with repect to CCA in terms of bias.]

@@

The simulator should choose their data generating mechanism(s) and estimand(s) in line with the study scope. Model-based data-generating mechanisms have the advantage of flexibility and precision, since data are generated from a probability distribution or statistical model and the true theoretical parameters can be known. A design-based approach is often used in situations where a probability distribution is not available, or where real-life data structures are of interest. A benefit of design-based simulation is the ability to use real-life observed data structures.

The comparative truth of the estimand(s) is determined by the choice of data generating mechanism and whether sampling variance is included in the simulation set-up, see Table \ref{table:dgm_truth}. If sampling variance cannot be omitted from the simulation scheme, multiple samples are drawn from the data generating mechanism (i.e., one sample per simulation repetition, a standard simulation set-up). If sampling variance is not of interest, a single complete data set can be obtained from the data generating mechanism (i.e., one sample for all simulation repetitions). This process is computationally convenient because only a single complete data set has to be considered during all of the simulations. We highly value the additional flexibility that such a simulation set-up offers. Especially in the case of data transformations it can be challenging to derive the true parametric references to evaluate against. Simply using a single generated complete set in which simulated missingness is induced and evaluating against that generated data set avoids a plethora of procedural problems and computational challenges.
% The simulator should first choose an appropriate data generating mechanism and then decide if sampling variance can be omitted from the simulation scheme to draw one or more samples from the data generating mechanism, respectively.
When deviating from the standard simulation workflow, conventional pooling rules for multiple imputation \citep[cf.][p. 76-77]{rubi87} do not apply. Instead, alternative pooling rules need to be used \citep{raghunathan2003multiple,vink14}.

\begin{table}[htb]
\begin{center}
\caption{Source of the comparative truth in simulation studies.}
\label{table:dgm_truth}
\begin{tabular}{lll}
\hline
               & With sampling variance      & Without sampling variance \\
               & (multiple samples drawn)    & (one sample drawn) \\
\hline  
Model-based simulation   & data-generating model         & single sample \\
Design-based simulation  & sufficiently large data set   & single sample \\
\hline
\end{tabular}
\end{center}
\end{table}

@@

[Missingness mechanisms:] 
First, one should always consider MCAR missingness, i.e. the scenario where the missing values are missing at random and the observed values are observed at random. % Under MCAR, the statistical properties of the observed data given the missing data are known and any imputation routine that cannot at least mimic the performance of the observed data inference, should be deemed inefficient in the scope of the simulation. 
Next, missing data should be induced conform a model that is dependent on the observed data. A straightforward technique for inducing different forms of univariate MAR missingness is described in \citet[][p. 63]{fimd} and a generalization to multivariate MAR missingness can be found in \citet{ampute}. If the missingness is to be induced in longitudinal data, autoregressive MAR models \citep[e.g. cf.][model 2 and model 3]{shara2015randomly} can be useful.
[Although often ignored, it can be argued that MNAR is the more likely mechanism for real-life missingness scenarios. It would be wise to include the mechanism in your simulation study, or perform sensitivity analyses.] An indication of the validity of the obtained inference, given that the assumed missingness mechanism is suspected to be invalid, may be obtained by performing sensitivity analysis \citep[see e.g.][part 5]{molenberghs2014handbook}.

[Missingness type:] 
Third, it is advisable to investigate varying shapes of MAR missingness to achieve a more realistic indication of the robustness of the imputation performance across the range of random missingness. The eefects of different forms of MAR mechanisms are described in \citet{scho18}.% Given the simulated data distributions, one random missingness model may be far more disastrous to the observed information than another model. This may influence the performance of some (but not necessarily all) imputation routines. For example, inference from hot-deck techniques such as predictive mean matching \citep{little1988missing, rubin1986statistical} may be more severely impacted by large amounts of one-tailed missingness than inference from parametric techniques.
It would be a shame to overlook results due to the focus on a single MAR mechanism.

@@

It could be convincing to demonstrate a method's applicability to real-world missing data problems. This can, for example, be achieved by obtaining and imputing a fully observed set of variables, wherein the missingness is mimicked from a similar, incomplete set. Alternatively, an incomplete set could be obtained and truth could be established by removing the incomplete cases from the data. However, the real-world missingness would then be omitted.
